### 2021-04-13 pldi21/discussion.tex
second round
updates in respond to shepherd

  The livelits mechanism as described in this paper operates only on expressions,
  but livelits might be useful for generating other sorts of terms, such as types,
  patterns, and entire modules. Prior work on literal macros has explored this \cite{TLMs}.
  % Support for type and module splices, too, would likely be useful.
  % With these, better support for reflecting on the provided type might allow for 
  % the development of more general livelits, e.g. a variant of \li{\$datatype} 
  % with individually typed columns, rather than only floating point data.

   The strict hygiene discipline has, we believe, substantial 
   benefits---programmers will inevitably encounter unfamiliar livelits, and 
   the reasoning principles that we enforce are likely to help them ``reason around''
   the situation. However, it may be useful in certain circumstances to 
   relax these, with the editor alerting the user to the unusual situation.

@@

\newpage
% \begin{itemize}
%   \item pattern matching
%   \item type splices
%   \item module splices
%   \item explicitly make bindings available in splices
%   \item better UI for closure provenance / connect with control flow / call stack better
%   \item deriving livelits from type definitions
%   \item bidirectional evaluation ala SnS
%   \item integration with structure editing more cleanly
%   \item \dots
%   \item collaborative editing?
%   \item side effects?
%   \item full screening livelits (or compositions thereof) as a way to create end-user workflows
%   \item big vision: authoring environment based on livelits
% \end{itemize}

>>>>

  The livelits mechanism as described in this paper operates only on expressions,
  but livelits might be useful for generating other sorts of terms, such as types,
  patterns, and entire modules. Prior work on literal macros has explored this \cite{TLMs}.
  (Note that Racket's visual macro system generates
  arbitrary syntax, so it can already be used in 
  this manner, albeit with no sort-specific semantic guarantees.)
  % Support for type and module splices, too, would likely be useful.
  % With these, better support for reflecting on the provided type might allow for 
  % the development of more general livelits, e.g. a variant of \li{\$datatype} 
  % with individually typed columns, rather than only floating point data.

   The strict binding discipline has, we believe, substantial 
   benefits---programmers will inevitably encounter unfamiliar livelits, and 
   the reasoning principles that we enforce are likely to help them ``reason around''
   the situation. However, it may be useful in certain circumstances to 
   relax these, with the editor alerting the user to the unusual situation.

@@

% \begin{itemize}
%   \item pattern matching
%   \item type splices
%   \item module splices
%   \item explicitly make bindings available in splices
%   \item better UI for closure provenance / connect with control flow / call stack better
%   \item deriving livelits from type definitions
%   \item bidirectional evaluation ala SnS
%   \item integration with structure editing more cleanly
%   \item \dots
%   \item collaborative editing?
%   \item side effects?
%   \item full screening livelits (or compositions thereof) as a way to create end-user workflows
%   \item big vision: authoring environment based on livelits
% \end{itemize}

\section*{Acknowledgements}

(Acknowledgements are being prepared and will be included in the version sent to the publisher.)

### 2021-04-13 pldi21/implementation.tex
updates in respond to shepherd

\subsection{Text Editor Integration}
Livelits do not require the use of a structure editor. 
We have also developed a 
proof-of-concept implementation of livelits in a textual program editor, Sketch-n-Sketch \cite{sns-pldi,sns-uist},
which recently added support for the necessary mechanisms \cite{DBLP:journals/pacmpl/LubinCOC20}.
Livelit interaction causes the serialized model in the text buffer to be changed, which updates the view.
This proof-of-concept is focused on this interaction so it is not at feature parity with the 
mechanism as described here or 
our main implementation, but it demonstrates that a syntax-recognizing text editor \cite{DBLP:journals/tosem/BallanceGV92,DBLP:conf/sde/HorganM84,interactive-visual-syntax} 
is sufficient to support livelits, albeit with some gaps in availability when 
there are syntax errors.
There are other systems that integrate visual syntax into an otherwise textual editor as well, notably Dr. Racket's recently introduced visual macro system \cite{interactive-visual-syntax}, discussed in the next section.

>>>>

\subsection{Text Editor Integration}
Livelits do not require the use of a structure editor. 
We have also developed a 
proof-of-concept implementation of livelit interaction in a textual program editor, Sketch-n-Sketch \cite{sns-pldi,sns-uist}.
Livelit interaction causes the serialized model in the text buffer to be changed, which updates the view.
This proof-of-concept is focused on this interaction loop so it is not at feature parity with the 
 main implementation, but it demonstrates that a syntax-recognizing text editor \cite{DBLP:journals/tosem/BallanceGV92,DBLP:conf/sde/HorganM84,interactive-visual-syntax} 
is sufficient to support livelits, albeit with some gaps in availability when 
there are syntax errors.
There are other systems that integrate visual syntax into an otherwise textual editor as well, notably Dr. Racket's recently introduced visual macro system \cite{interactive-visual-syntax}, discussed in the next section.

### 2021-04-13 pldi21/livelit-calculus.tex
updates in respond to shepherd

In order to communicate the semantics of livelits
independently
of the specifics of the Hazel environment and the web platform, 
we now specify a \emph{simply typed livelit calculus}.

Fig.~\ref{fig:syntax} specifies the syntax of the livelit calculus.
Programs are written as \emph{unexpanded expressions}, $\hat e$, which are \emph{expanded} to
\emph{external }(or \emph{expanded}) \emph{expressions}, $e$, before being \emph{elaborated}
to \emph{internal expressions}, $d$, for evaluation. All three sorts are classified
by the same types, $\tau$. We include partial functions, products, sums, and recursive
types, all in their standard form \cite{pfpl}, but this specific type structure is not critical.
Any language expressive
enough to encode its own abstract syntax 
would be a suitable basis for a livelit calculus.

We begin in Sec.~\ref{sec:external-and-internal-lang} with a terse
overview of the external and internal languages,
which are adapted from Hazelnut Live \cite{HazelnutLive}.
We then detail livelit expansion (Sec.~\ref{sec:calculus-expansion}) and 
liveness via closure collection (Sec.~\ref{sec:calculus-closure-collection}). 
% and parameterization (Sec.~\ref{sec:calculus-parameterization}).

@@

External expressions elaborate to internal expressions, $d$,
by the judgement $\elabs{\Gamma}{e}{d}{\tau}{\Delta}$.
The main purpose of elaboration is to initialize the substitution $\sigma$ on each hole closure, $\dehole{u}{\sigma}$, 
which takes the form $[d_i/x_i]_{i<n}$ and serves to capture
the substitutions that occur around the hole during evaluation. The key rule is:
\begin{mathpar}
\inferrule[Elab-Hole]{ }{
    \elabs{\Gamma}{\hehole{u}}{\dehole{u}{\idof{\Gamma}}}{\tau}{u :: \tau[\Gamma]}
}
\end{mathpar}
In this declarative formulation, the type $\tau$ is arbitrary (in Hazelnut Live, elaboration is specified algorithmically \cite{HazelnutLive}).
The substitution is initially the identity substitution, $\idof{\Gamma}$, i.e. the
substitution that maps each variable in $\Gamma$ to itself, because no substitutions have yet occurred. For example,
\[ \elabs{ }{(\hlam{x}{\hehole{u}})~\hnum{5}}{(\hlam{x}{\dehole{u}{[x/x]}})~{\hnum{5}}}{\tnat}{u :: \tnum[x : \tnat]} \]

>>>>

In order to specify the semantics of livelits
independently
of the specifics of the Hazel environment and the web platform, 
we now specify a \emph{simply typed livelit calculus}.

Fig.~\ref{fig:syntax} specifies the syntax of the livelit calculus.
Programs are written as \emph{unexpanded expressions}, $\hat e$, which are \emph{expanded} to
\emph{external }(or \emph{expanded}) \emph{expressions}, $e$, before being \emph{elaborated}
to \emph{internal expressions}, $d$, for evaluation. All three sorts are classified
by the same types, $\tau$.

We include partial functions, products, sums, and recursive
types, all in their standard form \cite{pfpl}, but this specific type structure is not critical.
Any language expressive
enough to encode its own abstract syntax 
would be a suitable basis.

% We begin in Sec.~\ref{sec:external-and-internal-lang} with a terse
% overview of the external and internal languages,
% which are adapted from Hazelnut Live \cite{HazelnutLive}.
% We then detail livelit expansion (Sec.~\ref{sec:calculus-expansion}) and 
% liveness via closure collection (Sec.~\ref{sec:calculus-closure-collection}). 
% and parameterization (Sec.~\ref{sec:calculus-parameterization}).

@@

External expressions elaborate to internal expressions, $d$,
according to the declarative elaboration judgement $\elabs{\Gamma}{e}{d}{\tau}{\Delta}$.
(Hazelnut Live specifies elaboration algorithmically \cite{HazelnutLive}.)
The main purpose of elaboration is to initialize the substitution $\sigma$ on each hole closure, $\dehole{u}{\sigma}$, 
which takes the form $[d_i/x_i]_{i<n}$ and serves to capture
the substitutions that occur around the hole during evaluation. The key rule for elaboration at type $\tau$ is:
\begin{mathpar}
\inferrule[Elab-Hole]{ }{
    \elabs{\Gamma}{\hehole{u}}{\dehole{u}{\idof{\Gamma}}}{\tau}{u :: \tau[\Gamma]}
}
\end{mathpar}
% In this declarative formulation, the type $\tau$ is arbitrary .
The substitution is initially the identity substitution, $\idof{\Gamma}$, i.e. the
substitution that maps each variable in $\Gamma$ to itself, because no substitutions have yet occurred. For example,
\[ \elabs{ }{(\hlam{x}{\hehole{u}})~\hnum{5}}{(\hlam{x}{\dehole{u}{[x/x]}})~{\hnum{5}}}{\tnat}{u :: \tnum[x : \tnat]} \]

### 2021-04-13 pldi21/livelit-definitions.tex
updates in respond to shepherd

\noindent
We now take the perspective of a livelit provider.
Fig.~\ref{fig:color-impl}
defines \li{\$color} from Fig.~\ref{fig:color}. We omit certain 
incidental details and use unimplemented syntactic sugar, including
Haskell-style \li{do} notation and quasiquotation, for presentation.

Livelit definitions are scoped and packaged like 
any other definition. They consist of a declaration and an implementation. 
Line 2 of Fig.~\ref{fig:color-impl} is \li{\$color}'s declaration,
which defines its name and its {expansion type}, \li{Color}, defined on Line 1. 
Livelit parameters would also appear here as shown in the declarations in Sec.~\ref{sec:expansion-typing}. 
The declaration is part of the client interface, 
as discussed in Sec.~\ref{sec:expansion-typing}-\ref{sec:parameterization}.

\subsection{Livelit Implementations}
The curly braces delimit the livelit's implementation, which is not intended to be seen 
by clients. 
Every livelit implementation must include all of the definitions shown in Fig.~\ref{fig:color-impl}, 
which we will discuss in turn below. 
Livelits are implemented using a variation on the functional model-view-update
architecture popularized by Elm \cite{ElmArchitecture}. We add a fourth component,
expansion generation. In addition, we use a simple monadic framework (\emph{a la} Haskell \cite{marlow2010haskell}) to provide the necessary  
interface between the livelit and the editor, while retaining
a pure functional programming model (monadic commands, like \li{new_splice} discussed below, are executed by the editor).

\subsubsection{Model}\label{sec:model}
The state of a livelit's GUI is determined by its model value. 
Line 3 of Fig.~\ref{fig:color-impl} specifies the corresponding \emph{model type},
here a labeled 4-tuple of \emph{splice references}, one for each of the four splices
that appear in the GUI in Fig.~\ref{fig:color}.
The model is how the GUI state is persisted in the syntax tree, so 
 the system requires that the model type supports automatic serialization
 (so no functions can appear in models).

@@

\subsubsection{Expansion}
\label{sec:expansion}
The ultimate purpose of a livelit is to fill the hole where it appears by generating an expansion,
i.e. an expression of the expansion type, here \li{Color}.
The \li{expand} function determines this expansion based on the model 
(which encodes the GUI state, and so usually contains information, like splice references
and widget states, 
that should not persist into run-time as part of the expansion).

The expansion can include the spliced expressions, 
but the system does not make the expression  
available directly as a value of type \li{Exp}. 
Instead, \li{expand} returns a \emph{parameterized expansion}, of type \li{Exp}, 
paired with a list of \li{SpliceRef}s (which can come from 
both splices and parameters as discussed above).
The {parameterized expansion}
 is an encoding (i.e. a value of type \li{Exp}) of a function 
 (here curried)
 that takes an argument for each listed \li{SpliceRef}. 
 That argument is of the corresponding splice 
type, which was provided when the splice was initialized. 
 The return type of the parameterized expansion is the expansion type.
 So here, the parameterized expansion for \li{\$color} must be a function of type 
 \li{Int -> Int -> Int -> Int -> Color}.

Note that Hazel does \emph{not} statically check the definition of \li{expand}
to ensure that the encoded parameterized expansion has this type. Instead, the parameterized expansion 
is \emph{validated} at each livelit invocation site, with errors reported to the client. 
A \emph{typed quotation} system as in, e.g., MetaOCaml \cite{DBLP:conf/flops/Kiselyov14}, could be adapted to allow for definition-site verification, as discussed in \cite{TLMs}, but 
note that   
the type of the quotation depends on the type of each splice in 
the splice list.

This parameterization strategy makes it simple to enforce the binding discipline described in Sec.~\ref{sec:hygiene}. 
The parameterized expansion 
can depend only on the \li{context}, whereas the splices are entered by the client and so they can 
depend only on the client site typing context. The use of function application ensures
that splices are capture avoiding as discussed in Sec.~\ref{sec:hygiene}. 
We consider this more formally next.

>>>>

\noindent
We will now take the perspective of a livelit provider.
Fig.~\ref{fig:color-impl}
defines \li{\$color} from Fig.~\ref{fig:color}, 
which is our prototypic example of a livelit 
definition. 
We omit certain  
incidental details and use unimplemented syntactic sugar, including
Haskell-style \li{do} notation \cite{marlow2010haskell} and quasiquotation \cite{bawden1999quasiquotation}, for presentation.

Livelit definitions are scoped and packaged like 
any other definition. Each \emph{definition} consists of a \emph{declaration} and an \emph{implementation}. 
In Hazel, a ``template'' declaration and implementation is generated 
as soon as the user types ``\li{livelit\ }''.

\subsection{Livelit Declarations} 
Line 2 of Fig.~\ref{fig:color-impl} is \li{\$color}'s declaration,
which defines its name and its {expansion type}, \li{Color}, defined on Line 1. 
Both are required. 
Livelit parameters can also optionally appear here as shown in the declarations in Sec.~\ref{sec:expansion-typing}. 
The declaration is part of the client interface, 
as discussed in Sec.~\ref{sec:expansion-typing}-\ref{sec:parameterization}.

\subsection{Livelit Implementations}
The curly braces delimit the livelit's implementation, which is not intended to be seen 
by clients. Each livelit implementation must define types \li{Model} and \li{Action}, 
values \li{init}, \li{view}, \li{update}, and \li{expand} with the signatures specified in Fig.~\ref{fig:color}, 
and a \li{context}, which is a listing of definition-site bindings that the livelit can make use of as described below. All of these are included in the generated template.
 These constitute a variation on the pure functional model-view-update
architecture popularized by Elm \cite{ElmArchitecture}. We add a fourth component,
expansion generation. In addition, we use a monadic framework (\emph{a la} Haskell \cite{marlow2010haskell}) to provide a pure   
 interface between the livelit and the editor: monadic commands, like \li{new_splice} discussed below, are executed by the editor,
and \li{do} notation provides syntactic sugar for monadic bind.

\subsubsection{Model}\label{sec:model}
The state of a livelit's GUI is determined by its model value. 
Line 3 of Fig.~\ref{fig:color-impl} specifies the corresponding \emph{model type},
here a labeled 4-tuple of \emph{splice references}, one for each of the four splices
that appear in the GUI in Fig.~\ref{fig:color}.
The model is how the GUI state is persisted in the syntax tree, so 
 the system requires that the model type supports automatic serialization
 (so functions cannot appear in models).

@@

\subsubsection{Expansion}
\label{sec:expansion}
The ultimate purpose of a livelit is to fill the hole where it appears by generating an expansion,
i.e. an expression of the expansion type, here \li{Color}.
The \li{expand} function determines the expansion given the model. 
The two are necessarily distinct: the model encodes the GUI state, 
and so usually tracks ephemeral widget state, like which tab is active in a tabbed GUI, 
as well as splice references, which refer only indirectly to the spliced expressions themselves, discussed below. 
This information is not needed at run-time as part of the expansion.
Only the model is persisted when a Hazel program is saved, because the expansion 
can be regenerated by calling \li{expand} whenever needed.

The expansion can include spliced expressions, 
but the system does not make these expressions  
available directly as values of type \li{Exp}. 
Instead, the expansion must treat splices parametrically. 
In particular, \li{expand} returns an encoded \emph{parameterized expansion}, of type \li{Exp}, 
paired with a list of \li{SpliceRef}s (which can come from 
both splices and parameters as discussed above).
The {parameterized expansion}
  is a function 
 (here curried)
 that takes an argument for each listed \li{SpliceRef}. 
 That argument is of the corresponding splice 
type, which was provided when the splice was initialized. 
 The return type of the parameterized expansion is the expansion type.
 So here, the parameterized expansion for \li{\$color} must be a function of type 
 \li{Int -> Int -> Int -> Int -> Color} because the splices 
 in the splice list were each of type \li{Int}.

This parameterization strategy makes enforcing the binding discipline described in Sec.~\ref{sec:hygiene}
straightforward. 
Context independence is maintained by allowing the parameterized expansion 
to depend only on bindings given in the explicit \li{context}, which 
the system maintains bound relative to the definition site. 
The splices can 
depend only on the client site typing context. The use of function application ensures
that splices are capture avoiding. 
% We consider this more formally next.

Note that Hazel does \emph{not} statically check the definition of \li{expand}
to ensure that the encoded parameterized expansion has the necessary type in the specified \li{context}. Instead, the parameterized expansion 
is only \emph{validated} at each livelit invocation site, with errors reported to the client. 
A \emph{typed quotation} system as in, e.g., MetaOCaml \cite{DBLP:conf/flops/Kiselyov14}, could be adapted to allow for definition-site verification, as discussed in \cite{TLMs}, but 
note that   
the type of the quotation depends on the type of each splice in 
the splice list, which has arbitrary length in general (e.g. the 
number of splices in the \li{\$dataframe} livelit 
depends on how many rows and columns the user has added).
